---
title: "Perceptron Linear Algebra"
output: html_notebook
---

## Introduction


### R Packages
Before we begin, we can go ahead and import the required R packages. If they are
not currently installed, that can be done first. The packages used in this 
project are:
- [data.table](https://cran.r-project.org/web/packages/data.table/index.html)
- [ggplot](https://ggplot2.tidyverse.org/)
- [plotly](https://plotly.com/r/)

```{r}
# Install if needed
if (!require(data.table)) install.packages('data.table')
if (!require(ggplot2)) install.packages('ggplot2')
if (!require(ggplot2)) install.packages('ggplot2')

library(data.table)
library(ggplot2)
library(plotly)
```

## Math Fundamentals of the Perceptron Algorithm
In order to understand the Perceptron algorithm, a basic knowledge of the
following mathematical principles must first be understood:
- Vectors (specifically, how to calculate the direction and norm)
- The Dot Product

I am going to assume that majority of people reading this article have prior 
knowledge of each of these fundamentals. If you do not, a great resource can be
found [here](https://www.khanacademy.org/math/precalculus/x9e81a4f98389efdf:vectors). I will also give a brief refresher to provide some context.

### Vectors

```{r}

```

### The Dot Product

```{r}

```

### Linear Separable Data
The definition of linearly separable data is pretty self-explanatory. At a high
level, for data to be linearly separable it simply means that there exists some
plane that can literally "separate" the data. For example, consider the
following plot:

```{r}
two_d_sample <- data.table(
  x = c(seq(1, 4, .1), seq(7, 10, .1)),
  y = c(runif(31, 1, 4), runif(31, 7, 10)),
  class_ = as.factor(c(rep(1, 31), rep(0, 31)))
)

plt1 <- ggplot(two_d_sample, aes(x=x, y=y, color=class_)) + 
    geom_point()

plt1
```

It is pretty obvious that you can draw some arbitrary line that separates the 
two different colored data points:

```{r}
plt1 + geom_abline(slope=-1.25, intercept=12)
```

Similarly, consider the following three-dimensional data:

```{r}
three_d_sample <- two_d_sample[, z := c(runif(31, 1, 4), runif(31, 7, 10))]

scene = list(
  camera = list(
    center = list(x = 0, y = 0, z = 0), 
    eye = list(x = 1.75, y = .75, z = .25)
    )
  )
plt2 <- plot_ly(data=three_d_sample, 
                x=~x, y=~y, z=~z, 
                color=~class_,
                type='scatter3d',
                mode='markers', 
                colors=c('darkred', 'blue')) %>%
  layout(scene=scene)

plt2
```

In this instance, a line would be insufficient when attempting to seperate the 
two data classes. Instead, a two-dimensional plane can be used:

```{r}

plt3 <- plot_ly(data=three_d_sample, 
                colors=c('darkred', 'gray', 'blue')) %>%
  add_trace(x=~seq(min(x), max(x), .1)*1.5, 
            y=~seq(min(y), max(y), .1)*1.5, 
            z=~matrix(6, nrow=62, ncol=62),
            type='surface') %>%
  add_trace(x=~x, y=~y, z=~z, 
                color=~class_,
                type='scatter3d',
                mode='markers') %>%
  layout(scene=append(scene, 
                      list(xaxis=list(title='x'),
                           yaxis=list(title='y'),
                           zaxis=list(title='z')))) %>% 
  hide_colorbar()

plt3
```

Of course, this gets a little harder to visualize - or even explain - 
as dimensionality continues to increase. As a result, a more generalized term
to describe this linear separation is used: a **hyperplane**.

### Hyperplanes
According to the [Wikipedia definition](https://en.wikipedia.org/wiki/Hyperplane): 
"In geometry, a hyperplane is a subspace whose dimension is one less than 
that of its ambient space." This makes sense. Looking at our first example, the
deminsionality of the data is two, and the line drawn to separate the data is
one-dimensional. In the second example, the data is three-dimensional and the
separation is two.

## Perceptron Algorithm

### Generate Data

```{r}

```

### Weights

```{r}

```

### Iterate

```{r}

```


## Multi-Layer Perceptron

```{r}

```







